taxadb: A High-Performance Local Taxonomic Database Interface

### 2019-03-04 paper/manuscript.Rmd
WIP

1. A problem already solved  including ITIS, NCBI, CoL, GBIF, WikiData, FishBase, and The Plant List

@@

@@

`get_ids` by default looks only for exact matches.  We can get more matches by cleaning the names first.  The `clean_names()` function provides several simple string manipulations to tidy names to improve the probability of a match: missing species (specific eptithets) such as `Accipiter sp.` drop the `sp.`, allowing matches against Genus names, and intraspecific epithets such as `Colaptes auratus cafer` are dropped to binomial names, `Colaptes auratus`.  These transformations may not be appropriate for certain use cases, and should be used with care, and only once as many untransformed names as possible have already been matched.

In this example, applying this transformation on the unmatched names we can bring matches against OTT identifiers up from 665 to 740, missing only ten names:

```{r}
ott_ids <- bbs %>% 
  mutate(names = clean_names(species)) %>% 
  pull(names) %>% 
  ids("ott")

unmatched <- ott_ids %>% 
  filter(is.na(acceptedNameUsageID)) %>% 
  select(input, sort)
```

@@

```{r}
unmatched %>% mutate(iucn = get_ids(input, "iucn", "prefix"))
```

@@

```{r}
syn <- unmatched$input %>% synonyms("iucn")
```

```{r}
x <- syn %>% 
  mutate(ott_id = get_ids(synonym, "ott", "prefix")) %>% 
  select(input, ott_id) %>%
  na.omit()

syn$synonym %>% get_ids("ott", "prefix")

descendants(id = x$ott_id, provider = "ott") %>% select(scientificName, taxonomicStatus, acceptedNameUsageID)
```

@@

```{r}
syn$synonym %>% na.omit() %>% ids("ott")
```

>>>>

1. A problem already "solved" including ITIS, NCBI, CoL, GBIF, WikiData, FishBase, and The Plant List

@@

Outline strategy for name resolution.

ID resolution including synonym mapping.

@@

`get_ids` looks only for exact matches.  Certain names may not uniquely resolve to a single ID, such as a name which is a recognized synonym for two separate taxonomic groups (such as may happen if a species is split.)

We can get more matches by cleaning the names first.  The `clean_names()` function provides several simple string manipulations to tidy names to improve the probability of a match: missing species (specific eptithets) such as `Accipiter sp.` drop the `sp.`, allowing matches against Genus names, and intraspecific epithets such as `Colaptes auratus cafer` are dropped to binomial names, `Colaptes auratus`.  These transformations may not be appropriate for certain use cases, and should be used with care.

Observe that truncating names would result in a handful of subspecies being resolved only to the species level identifiers

```{r}
bbs_cleaned_ids <- 
  bbs_ids %>% 
  select(species, ott) %>% 
  mutate(ids_post_clean = 
           get_ids(clean_names(species, binomial_only = TRUE), 
                   "ott", "prefix"))
bbs_cleaned_ids %>% na.omit() %>% filter(ott != ids_post_clean)
x <- bbs_cleaned_ids %>% na.omit() %>% filter(ott != ids_post_clean)

x
descendants(id = x$ids_post_clean, provider = "ott")
```

```{r}
bbs_ids %>% filter(is.na(ott) & !is.na(ids_post_clean))
```

```{r}
pick <- function(A, B){ A[is.na(A)] <- B[is.na(A)]; A }

ott_ids <- bbs_cleaned_ids %>% mutate(ids = pick(ott, ids_post_clean))
```

```{r}
unmatched <- ott_ids %>% 
  filter(is.na(ids)) %>% 
  select(species, ids) %>%
  mutate(species = clean_names(species))

unmatched
```

@@

```{r}

unmatched %>% mutate(iucn = get_ids(species, "iucn", "prefix"))
```

@@

```{r}
syn <- unmatched$species %>% synonyms("iucn")
syn
```

```{r}
x <- syn %>% 
  mutate(ott_id = get_ids(synonym, "ott", "prefix")) %>% 
  select(input, ott_id) %>%
  na.omit()
x
```

```{r}
syn$synonym %>% get_ids("ott", "prefix")
```

```{r}
descendants(id = x$ott_id, provider = "ott") %>% select(scientificName, taxonomicStatus, acceptedNameUsageID)
```

@@

```{r}
syn$synonym %>% na.omit() %>% ids("ott") %>% select(input, scientificName, acceptedNameUsageID, taxonomicStatus)
```
